The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity

### 2023-06-07 paper/markdown/rebuttal_letter.Rmd

<span custom-style = 'reviewer'>__Results:__ Page 9 lines 24-41; This paragraph should be in the Methods instead of the Results. Furthermore, the rationale is missing: Why was INCOV used as a comparison or additional study to your own data? In addition, information on the measurement /measurement system (for INCOV) should be provided. Were the measuring probes used in SIMMUN and INCOV comparable?</span>

As suggested, we now list the variables of interest in __Methods__ instead of __Results__. 
Furthermore, we describe the rationale for using the INCOV dataset in the last paragraph of __Introduction__ and the __Analysis endpoints__ subsection of __Methods__. 
Information on measurement of proteins and metabolites is provided in __Methods/Procedures__ and __Supplementary Methods__. Unfortunately, the incompatibility of designs and datasets of these two cohorts hindered us from developing a comprehensive model in one cohort and direct validation in the outer (canonical training - test setup). 
We list this issue in __Limitations__.

@@

We provide the time intervals for the acute, sub-acute and recovery timepoints in the INCOV cohort in the revised manuscript (__Methods__, __Results__, __Figures__).

## Issue 20

<span custom-style = 'reviewer'>__Results:__ What does the term "virtually age-independent" mean? I would appreciate it if concrete numbers are stated here instead.</span>

We have removed all imprecise wording while rewriting the __Results__ section. 
Thank you for pointing this out.

@@

This part of the manuscript was removed during revision. 
Following your suggestion, we keep details on procedures and statistical analysis in the __Methods__ section.

## Issue 22

<span custom-style = 'reviewer'>__Results:__ Page 15 line 28: standard inflammatory cytokines should be named more specifically.</span>

We are sorry for this unclarity. 
In the revised text, we name inflammatory markers (SIMMUN: NEO, NLR, INCOV: IL6, IL10, TNF and IFNG) more explicitly.

@@

## Issue 23

<span custom-style = 'reviewer'>__Discussion:__  The discussion of the study results is rather shallow: Results should be better integrated into existing evidence. Furthermore, there is no discussion of the results of depressive symptoms (HADS).</span>

We restructured and improved __Discussion__. 
Large parts of __Discussion__ were reorganized and rewritten as suggested by Reviewer 2 (__Issue 40__). 
We hope that you agree with the structure proposed by the other Reviewer which we have now followed. 
We have also included new references to back out results. 
The __Limitations__ section was also expanded and improved.

@@

## Issue 24

<span custom-style = 'reviewer'>__Discussion:__ The statement that "BH4 is also important for nitric oxide synthesis and therefore involved in oxidative stress" is not correct. (Page 16) BH4 is an essential co-factor of NO-synthases that produce NO with its function as a vasodilator and additional anti-pathogenic effects. Only if BH4 is missing, there is an uncoupling of NOS with subsequently increased reactive oxygen/nitrogen species. Thus, your "therefore involved in oxidative stress" is misleading in normal physiological states.</span>

We have removed most of the discussion on oxidative stress from the revised text since it is likely not so important for interpretation of our data. 
Instead we focus more on BH4 as the co-factor for synthesis of indolamine and catecholamine neurotransmitters [@Geisler2013; @Capuron2011; @Neurauter2008; @Vancassel2018].

@@

We apologize, this is really misleading and was reworded. 
We meant levels of inflammatory markers in SARS-CoV-2-uninfected individuals, which we state now explicitly.

## Issue 26

<span custom-style = 'reviewer'>__Highlights:__ In the Highlights, the results and perhaps the conclusion should be emphasized more than the theoretical background (Highlights 1 and 2) or the methodology (Highlights 3 and 4).</span>

The __Highlights__ were adapted according to your suggestions.

@@

* _Page 6 line 11: The hyphen is missing at "inflammatory"_ - does not apply, the section was rewritten

* _Page 8: lines 55-56: In the participants who were tested…_ - rewritten

* _Page 8 line 58: timepoint in [the?] the course sounds strange (and the article is missing), may-be better time between infection and assessment._ - changed to 'timepoints of SARS-CoV-2 infection' and 'days after diagnosis via PCR'

* _Page 9 line 52: What are nadir TRP concentrations?_ - this was changed to 'minimum'

@@

* _Page 12 line 24: "on the other hand" without "on the one hand"_ - we removed this colloquial phrase and other similar wording from the revised text

* _Page 13: line 4: a comma is missing_ - rewritten

* _Page 14 line 14: […] via their bidirectional influence_ - rewritten

* _Page 14 line 34: please delete the term "larger"_ - done as suggested

@@

We apologize for the typo, it was corrected to 'BIC'.

## Issue 30

<span custom-style = 'reviewer'>__Supplementary Figure S2:__ It is not clear why fitted generalized adaptive models are presented as graphs, as the associated statistic (Spearman's rho) assesses monotone relationships.</span>

We concur with the Reviewer's view, that visualization of trends for monotone associations is problematic. 
Due to profound changes in the structure of the revised manuscript, we do not use scatter plots to illustrate monotone correlations between non-normally distributed variables any more. 
Correlation analysis results in the INCOV cohort are now presented as simple bubble correlograms or visualized as force-directed graphs [@Briatte2021; @Csardi2006].

@@

# Reviewer 2

## Issue 31

<span custom-style = 'reviewer'>In this paper, the authors show the relationship between alterations in markers of inflammation, aminoacid precursors, and symptoms of depression, anxiety, and stress in individuals COVID-19-positive or not. 
They relate the results of these observations to the results obtained in another immune and metabolic study carried out at different time points of the COVID-19 trajectory and already published by a group of different investigators.
The methodology is excellent as can be expected from a research team specialized in the study of inflammation and its impact on the kynurenine metabolism pathway. 
The results are original and sufficiently important to be published. 
The main problem with the current manuscript is the confusion coming from the mode of presentation that mixes up in the results section data from the two studies. 
The discussion is circumvoluted and difficult to follow.</span>

We thank the Reviewer for appreciation of out text and the constructive feedback. 
Weaknesses in structure of the analyses were already pointed out by the Reviewer 1. 
We addressed them by employing the multi-parameter modeling results for identification of the most vital explanatory factors that affect neurotransmitter metabolism and subsequent more specific correlation analyses and statistical testing with those relevant predictors. 
To make the analysis of the two investigated cohorts easier to follow, we separated the results obtained with the SIMMUN and INCOV cohort. 
In more detail, we employ the SIMMUN cohort as an ‘exploratory’ cohort to identify factors affecting TRP, KYN, KYN/TRP, PHE and PHE/TYR from a wider palette of demographic, clinical, psychometric, inflammatory and SARS-CoV-2 associated variables. Subsequently, we attempted to validate the most important ones, age, inflammation and SARS-CoV-2 infection, in the INCOV cohort. Finally, we re-wrote __Discussion__ with a special focus on the crosstalk between inflammation and neurotransmitter biosynthesis and, in part controversial, relevance of systemic neurotransmitter availability for the central nervous system and mental health.

@@

Thanks for the careful assessment. 
We paid special attention to the grammar and spelling of the revised text.

## Issue 33

<span custom-style = 'reviewer'>The text would benefit from further editing by a specialized scientific English editing service</span>

The first author of the manuscript has extensive experience in scientific English. 
We are assuming that some inconsistencies in the first version of the manuscript arose from working on the manuscript in a large study team. 
The first author has now carefully reworked the text and we hope that this has now lead to a more consistent style and better quality.

@@

## Issue 34

<span custom-style = 'reviewer'>Line 7: by collectives you probably mean cohorts or population samples.</span>

Yes, in the manuscript  text the term ‘collective’ had the same meaning as ‘cohort’. 
We removed the word 'collective' since this obviously was not easy to understand and use consistently 'cohort' instead. 
For practical reasons, we avoid the wording ‘population sample’, which may suggest a random and hence representative draw from the general population. 
In fact neither the SIMMUN cohort enriched in psychiatric patients nor the INCOV cohort enriched in hospitalized COVID-19 patients pose a representative sample of the general population during the pandemic (see __Limitations__).

@@

## Issue 35

<span custom-style = 'reviewer'>The introduction ends with the statement that the study was conducted in an exploratory manner. Was that the objective of the present study in the absence of any specific hypothesis?</span>

As also pointed out by Reviewer 1, the last paragraph of __Introduction__ was not written clearly enough in the initial manuscript version. 
In the final paragraph of __Introduction__, we present now the ‘gap in the literature’, which is the still incompletely characterized effect of SARS-CoV-2 infection on metabolic pathways determining systemic availability of serotonin and dopamine. 
We also state the aim of our analysis, which was characterization of factors affecting the serotonin, kynurenine and catecholamine pathway activity during SARS-CoV-2 infection and recovery. 
Our study did  not have a fully exploratory character, since the candidate explanatory factors: age, sex, body mass index, mental and somatic conditions, smoking and alcohol, depression, anxiety, stress and inflammation have already been described to affect serotonin and dopamine metabolism in previous research, also from our own research consortium [@Hufner2019; @Hufner2020; @Hufner2021; @Hufner2015].

@@

We agree that the term 'parameter' may be misleading, especially in context of modeling and its results. 
We changed it to 'variable' throughout the revised text.

## Issue 37

<span custom-style = 'reviewer'>Legend of Figures: the use of the term decay is inappropriate, you probably mean metabolites</span>

This was reworded in the revised version, we utilize 'catabolism', 'breakdown' or 'conversion' instead.

@@

<span custom-style = 'reviewer'>The results are difficult to read as you mix up the two cohorts both in the text and in the figures. It would be preferable to first present the results of your own cohort, the SIMMUN cohort, and then test the generality of the results by assessing whether the same variations are found in the INCOV cohort with the limitation that this can only be done for the relationship between immune factors and neurotransmitter precursors.</span>

We agree and rearranged the complete __Results__ section. 
Please refer to __Issues 1__ and __31__ for details.

## Issue 39

<span custom-style = 'reviewer'> You utilize the terms mental disease, mental stress, mental symptoms, etc. Be more specific and replace these terms by what you measured, i.e., HADS-anxiety, HADS-depression, and PSS. 
You mention somatic symptoms in the discussion but what they refer to remains obscure. 
You also refer to somatic diseases. 
Pathologists would rather speak of physical disorders as opposed to psychiatric disorders.</span>

Thank you for your suggestions. 
We have specified the terms according to your suggestions throughout the revised manuscript. 
We removed the sentence from __Discussion__ referring to somatic symptoms, since those were not systematically surveyed in the SIMMUN cohort. 
Following your suggestion, we term ‘somatic conditions’ ‘physical disorders’ consequently in the revised text.

@@

## Issue 40

<span custom-style = 'reviewer'>The discussion is convoluted and needs to be simplified. 
It would be easier to follow if you would first discuss the significance of the association between inflammation and neurotransmitter precursors and then the association between these two sets of factors and what you call mental health measured by HADS and PSS, first in the immune -> HDAS symptomatology or kynurenine -> HADS symptomatology and then in the stress -> immune direction, remembering that in the case of your own cohort, the cross-sectional nature of your study does not allow you to get further than describing associations.</span>

We thank the Reviewer for this specific suggestion on the discussion section which we have gratefully taken up and rewritten large parts of __Discussion__ according to the following scheme:

* A brief generalizing summary of the results obtained with both cohorts and discussion of discrepant findings

* Effects of inflammation and infection, including SARS-CoV-2, on the serotonin and kynurenine pathways, and systemic levels of serotonin

* Effects of inflammation and infection on the catecholamine pathway and systemic availability of dopamine

* Mutual effects of mental health-related factors (e.g. depression, anxiety and mental stress) on activity of the serotonin, kynurenine and catecholamine pathways, and systemic serotonin and dopamine availability. Here we also underline that evidence for effects of peripheral serotonin, kynurenine and catecholamine pathways is limited and based in many cases on clinical association of psychiatric disorders with blood levels of the neurotransmitters, their precursors or competitor pathway products.

@@

## Issue 41

<span custom-style = 'reviewer'>Page 15: the study by Benedetti is at best suggestive but does not demonstrate anything.</span>

We removed this reference from __Discussion__ of the revised manuscript.

@@

We thank the Reviewer for appreciation of our manuscript and the valuable feedback.

## Issue 43

<span custom-style = 'reviewer'>In the abstract, the authors introduce ...markers of serotonin availability (TRP, KYN, KYN/TRP ratio)....(line 24). 
This statement is not accurate, as Serotonin is formed from TRP, but there is no KYN in its catabolic pathway. 
Only if the catabolism of TRP moves more to the kynurenine pathway, less TRP goes to the serotonin pathway, and the availability of serotonin is reduced. 
This argument is similar to the line 44.</span>

We absolutely agree with the Reviewer. 
To clarify that, we use consistently the terms 'serotonin pathway', 'kynurenine pathway' and 'catecholamine pathway' in the revised manuscript.

@@

## Issue 44

<span custom-style = 'reviewer'>The principle point is that it would help the reader to orient himself in such a large amount of data if the authors completed the manuscript with a design scheme of the INCOV and SIMMUN study</span>

Thank you for the suggestions. 
As described in __Issues 1__ and __38__, we present the results in cohort-wise manner, for the sake of better readability. 
We also included the analysis strategy scheme in __Supplementary Figure S1__.

@@

## Issue 45

<span custom-style = 'reviewer'>In line 16 on p. 6 are introducing breakdown products as KYN, TRP, KYN/TRP ratio ..... It would be more precise to introduce them in the order they are formed: TRP, KYN, KYN/TRP ratio ....pls, correct everywhere in the manuscript similar p. 9, line 29 ..... For example, correct position is given on the line 52 of p.11</span>

Thank you, we made the changes you suggest throughout the revised manuscript.

@@

## Issue 46

<span custom-style = 'reviewer'>p. 14, line 34 : a large meta-analyses ... include more citations</span>

We have improved this part of __Discussion__ and now report citations [@Hunt2020; @Maes2012; @Brown2021; @Marx2020;  @Fellendorf2022; @Moncrieff2022]] on the pros and cons of KYN pathway involvement in depression in the text and the __Limitations__ section.

@@

This was re-written as suggested.

## Issue 48

<span custom-style = 'reviewer'>p. 15, line 62 - where ???</span>

The '???' was likely a typo in a figure reference in the manuscript markdown document. 
We have paid special attention to the proper syntax in the revised text.

@@

This was re-written as suggested.

## Issue 50

<span custom-style = 'reviewer'> the given data do not match mathematically - check or explain why</span>

We apologize for the typos concerning numbers in the study inclusion scheme. 
Please refer to the correct numbers in the revised __Supplementary Figure S1__.
